Calibrating seed-based alignment heuristics with Sesame

### 2020-03-16 manuscript.tex
Simplify skip-n transfer matrix (in appendix)

Reads consist of only two kinds of objects: the segments terminated by
$\Downarrow$ (or $\Downarrow$-segments for short) and the tails, so the
dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$-segment
can be followed by another $\Downarrow$-segment or by the tail. The tail
cannot be followed by anything. The expression of the transfer matrix
$M_0(z)$ is thus
\begin{equation*}
\begin{blockarray}{ccc}
   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
\begin{block}{c[cc]}
\scriptstyle{\Downarrow} & \big( 1+qz+\ldots+(qz)^{\gamma-1}
\big) pz  & 1+qz+\ldots+(qz)^{\gamma-1} \\
\scriptstyle{|} & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

In the representation above, the different types of segments are
identified by their terminator, indicated in the margins for clarity.

@@

Those segments follow a $\Downarrow_i$-segment, so they start $i$
nucleotides in front of the next non-skipped position. The shortest
following segment that ends $j$ nucleotides before a non-skipped position
has length $\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is
equivalent to defining $x$ as the number of $\square$ symbols of the
shortest $\Downarrow_j$-segment, with $x = i-j-1 \modulo{n+1}$ and $\ell_0
= x+1$.

The shortest $\Downarrow_j$-segment consists of $x$ symbols $\square$
followed by the $\Downarrow_j$ terminator, so it has weighted generating
function $(qz)^xpz$. The other segments have a multiple of $n+1$
additional $\square$ symbols so their weighted generating functions are
$(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots, (qz)^{x+m(n+1)}pz$. Summing
over those cases, we finally obtain
\begin{eqnarray}
H_{i,j}(z) &=& (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor.
\end{eqnarray}

Denote $J_i(z)$ the weighted generating function of tails that can follow
a $\Downarrow_i$-segment. There are $\gamma+i$ of them, each consisting of
0 to $\gamma+i-1$ symbols $\square$ followed by the special $|$
terminator for the end of the read. The $|$ has size 0 so its weighted
generating function is 1. Once again summing over the different cases, we
obtain
\begin{eqnarray}
J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
\end{eqnarray}

Finally, the expression of $M_n(z)$ the transfer matrix of reads in the
skip-$n$ alphabet is
\begin{equation}
\label{eq_matrix_Mn}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation}

@@

But we can do even better: a read with $s+1$ segments contains $s$ errors,
so all the entries of $M_n(z)^{s+1}$ are dominated by $p^s$ and they
rapidly converge to $0$ as $s$ increases. Instead of computing the matrix
$M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$, we can interrupt the
summation after a certain power of $M_n(z)$ because the terms become
negligible.

@@

@@

\section{Off-target skip seeds}

@@

As in the case of the skip encoding, a fair amount of bookkeeping is
required to specify the weighted generating functions of interest.
We reuse the quantities $a+b+c+d=1$ and give them the same meaning as in
section~\ref{sec:dual}. For notational convenience, we define
\begin{eqnarray*}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& dz \cdot (az)^x \cdot \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big), \\
  &\quad& \text{where } x = n-j \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor. \\
U_{i,j}(z) &=& dz \cdot (az)^x \cdot
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = n-i+1 \modulo{n+1}, \\
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& dz \cdot (az)^x \cdot
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray*}

@@

>>>>

Reads consist of only two kinds of objects: the segments terminated by
$\Downarrow$ (or $\Downarrow$-segments for short) and the tails, so the
dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$-segment
can be followed by another $\Downarrow$-segment or by the tail. The tail
cannot be followed by anything. The expression of the transfer matrix
$M_0(z)$ is thus
\begin{equation*}
\begin{blockarray}{ccc}
   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
\begin{block}{c[cc]}
\scriptstyle{\Downarrow} & \big( 1+qz+\ldots+(qz)^{\gamma-1}
\big) pz  & 1+qz+\ldots+(qz)^{\gamma-1} \\
\scriptstyle{|} & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
where $p$ is the error rate of the sequencer, $q=1-p$ and $\gamma$ is the
minimum seed length. In the representation above, the different types of
segments are identified by their terminator, indicated in the margins for
clarity.

@@

Those segments follow a $\Downarrow_i$-segment, so they start $i$
nucleotides in front of the next non-skipped position. The shortest
following segment that ends $j$ nucleotides before a non-skipped position
has length $\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is
equivalent to defining $\ell_0$ as $x+1$ where $x$ is the number of
$\square$ symbols of the shortest $\Downarrow_j$-segment, \textit{i.e.} $x
= i-j-1 \modulo{n+1}$.

With these notations, he shortest $\Downarrow_j$-segment consists of $x$
symbols $\square$ followed by the $\Downarrow_j$ terminator, so it has
weighted generating function $(qz)^xpz$. The other segments have a
multiple of $n+1$ extra $\square$ symbols so their weighted
generating functions are $(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots,
(qz)^{x+m(n+1)}pz$. Summing over those cases, we finally obtain
\begin{eqnarray}
\label{eq_H}
H_{i,j}(z) &=& (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor.
\end{eqnarray}

Denote $J_i(z)$ the weighted generating function of tail segments that can
follow a $\Downarrow_i$-segment. There are $\gamma+i$ such tails, each
consisting of 0 to $\gamma+i-1$ symbols $\square$ followed by the special
$|$ terminator for the end of the read. The $|$ has size 0 so its weighted
generating function is 1. Once again summing over the different cases, we
obtain
\begin{equation}
\label{eq_J}
J_i(z) = 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
\end{equation}

Finally, the expression of $M_n(z)$ the transfer matrix of reads in the
skip-$n$ alphabet is
\begin{equation}
\label{eq_matrix_Mn}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation}
where $p$ is the error rate of the sequencer, $q=1-p$, $n$ is the number of
skipped nucleotides between potential seeds, $\gamma$ is the minimum seed
length, and polynomials $H_{i,j}(z)$ and $J_i(z)$ are defined in
expressions (\ref{eq_H}) and (\ref{eq_J}).

In appendix~\ref{app_Mn}, we present a transfer matrix with a simpler
expression that will prove useful in section~\ref{sec_offskip}. The
version of appendix~\ref{app_Mn} is simpler, but the expression above has
some advantages that will be explained below.

@@

But we can do even better: a read with $s+1$ segments contains $s$ errors,
so the top right entry of $M_n(z)^{s+1}$ is the weighted generating
function of reads with $s$ errors that have no seed of minimum size
$\gamma$. Computing the partial sum $M_n(z) + M_n(z)^2 + \ldots +
M_n(z)^s$ instead of $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$
corresponds to neglecting reads with $s$ or more errors. For $s$
sufficiently large, such reads are exceedingly rare so we can obtain
accurate estimates without computing all the powers of $M_n(z)$ up to
order $k$.

@@

If we used $M_n^*(z)$the transfer matrix of appendix~\ref{app_Mn}, the top
right entry of $M_n(z)^{s+1}$ is not the weighted generating function of
reads with $s$ errors, so (\ref{eq:bound}) is not an upper bound for the
neglected terms of the sum. As a consequence, one would have to compute
more terms in the partial sum $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{s+1}$
to reach the same accuracy. The transfer matrix shown in expression
(\ref{eq_matrix_Mn}) is not the simplest, but it has the benefit of
requiring fewer iterations.

@@

\section{Off-target skip seeds}
\label{sec_offskip}

@@

As in the case of the skip encoding, a fair amount of bookkeeping is
required to specify the weighted generating functions of interest.
We reuse the quantities $a+b+c+d=1$ and give them the same meaning as in
section~\ref{sec:dual}. For notational convenience, we define
\begin{eqnarray}
\label{eq_N}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& dz \cdot (az)^x \cdot \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big), \\
  &\quad& \text{where } x = n-j \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor. \\
U_{i,j}(z) &=& dz \cdot (az)^x \cdot
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = n-i+1 \modulo{n+1}, \\
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& dz \cdot (az)^x \cdot
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray}

@@

\subsection{An alternative construction of $M_n(z)$}
\label{app_Mn}

The alphabets described here are constructed so that there is exactly one
way to decompose reads in symbols or in segments. For a given alphabet
there exists only one transfer matrix that describes the reads that have
no seed of minimum size $\gamma$. However, if we choose a different
alphabet, we can obtain a different transfer matrix describing exactly the
same reads.

In section~\ref{sec:example_skip}, we decomposed the reads in
the so-called skip-$n$ alphabet $\mathcal{A}_n = \{\square, |,
\Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n \}$, from which we
derived the transfer matrix $M_n(z)$ shown in expression
(\ref{eq_matrix_Mn}). We now introduce the alternative skip-$n$ alphabet
$\mathcal{A}^*_n = \{\square, *, |, \Downarrow_0, \Downarrow_1, \ldots,
\Downarrow_n \}$, where we add the symbol $*$.

The $*$ symbol stands for the nucleotides between a $\Downarrow_j$
terminator ($0 \leq j \leq n)$ and the following non-skipped position.
Recall that seed can only start at non-skipped positions occurring every
$n+1$ nucleotides, and that the $\Downarrow_j$ indicates a sequencing
error $j$ nucleotides before a non-skipped position.
Figure~\ref{fig_alt_skip} shows the read of Figure~\ref{fig:skip} in the
alternative skip-3 alphabet. The $*$ symbols appear after the
$\Downarrow_1$ and $\Downarrow_3$ terminators. They are not found after a
$\Downarrow_0$ terminator because there are no nucleotides between this
terminator and the following non-skipped position.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{alternative_sketch_skip.pdf}
\caption{\textbf{The alternative skip encoding.}
The read of Figure~\ref{fig:skip} is represented in the alternative skip-3
alphabet. The only difference is that the nucleotides between a
$\Downarrow_j$ terminator ($0 \leq j \leq n$) and the following
non-skipped position are replaced by the $*$ symbol.}
\label{fig_alt_skip}
\end{figure}

The match status of the $*$ nucleotides is irrelevant. Indeed, after a
$\Downarrow_j$ symbol one has to ``wait''until the next potential seed can
start, $j$ nucleotides further down. The weighted generating function of a
$*$ nucleotide is thus $z$, as it stands of any nucleotide.

A sequence of $*$ symbols can end in only two ways: either with the end of
the read (as in the case of Figure~\ref{fig_alt_skip}), or with a
non-skipped position. If the end of the read comes first, the sequence of
$*$ symbols is terminated by the usual $|$ terminator. Otherwise, the
sequence has no terminator but it puts the read in exactly the same state
as after a $\Downarrow_0$ terminator.

This means that in the alternative transfer matrix, the entries associated
to $\Downarrow_j$ ($1 \leq j \leq n)$ simply ``fast forward'' to the
entries associated with $\Downarrow_0$ and $|$. It takes exactly $j$
nucleotides of any kind to go from a $\Downarrow_j$ symbol to the next
non-skipped position, with weighted generating function $z^j$. If the end
of the read occurs before the next non-skipped position, the
$\Downarrow_j$ will be followed by a tail segment of up to $j-1$ symbols
$*$, with weighted generating expression $1 + z + \ldots + z^{j-1}$. The
segments following the $\Downarrow_0$ terminator are unchanged compared to
the original expression of the transfer matrix.

Finally, the expression of  the transfer matrix $M_n^*(z)$ in the
alternative skip-$n$ alphabet is
\begin{equation}
\label{eq_alt_Mn}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & z & 0 & \ldots & 0 & N_0(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & z^n & 0 & \ldots & 0 & N_{n-1}(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation}
where $p$ is the error rate of the sequencer, $q=1-p$, where $n$ is the
number of skipped nucleotides between potential seeds, where $\gamma$ is
the minimum seed length, and where
\begin{gather}
\tag{\ref{eq_H}}
H_{0,j}(z) = (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  \text{with } x = n-j \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
\tag{\ref{eq_J}}
J_0(z) = 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma-1}, \\
\tag{\ref{eq_N}}
N_i(z) = 1+z+\ldots+z^i.
\end{gather}

Observe that this expression is simpler than (\ref{eq_matrix_Mn}). It
could have been used in section \ref{sec:example_skip} because it is also
easier to derive, but the burden of computations with expression
(\ref{eq_alt_Mn}) is heavier. The reason is that the top right entry of
$M^*_n(z)^{s+1}$ is not the weighted generating function of reads with $s$
sequencing errors ($s \geq 1$), so we cannot use the probability that the
reads contains more than $s$ errors as a bound on the neglected terms. As
a consequence, we would need to perform more iterations in order to
ascertain that the estimates are accurate to within chosen $\varepsilon$.
